---
title: "Danish Kings"
author: "August Harden"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator: 

After having opened up my own created file (called "Kongerække i Danmark") to view my danish monarchs in TextEdit, I can see that every data in the dataset is seperated by semicolons shown with the symbol: ";".


2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
library(tidyverse)
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/Kongerække i Danmark.csv")

kings2 <- read_csv("data/Kongerække i Danmark.csv")

kings3 <- read.csv2("data/Kongerække i Danmark.csv")

kings4 <- read_csv2( "data/Kongerække i Danmark.csv")

kings1
kings2
kings3
kings4

class(kings1)
class(kings2)
class(kings3)
class(kings4)

```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object.

When typing in all the different `kings` (1, 2, 3, and 4) loaded with my own dataset of danish monarchs, the function that shows to be a `tidyverse` function is `kings4`, and `kings2`.


2. What is the result of running `class()` on the `kings` object created with a tidyverse function?

When I use the `tidyverse` function (in my case `kings4`) I type `class(kings4)`, and get this in the output:
[1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 


3. How many columns does the object have when created with these different functions?

`kings4` and `kings3` have 5 coloumns each, while `kings2` and `kings1` just have 1 coloumn each.


4. Show the dataset so that we can see how R interprets each column.

I type `glimpse(kings)` to see how R interprets each coloumn.


```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv2( "data/Kongerække i Danmark.csv")

class(kings)


glimpse(kings)


head(kings)


```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
# If my data should be messy, I start by tidying it:
library(tidyverse)

# Now I will have to remove NAs from the dataset:
end_of_reign <- kings$`End of reign`
start_of_reign <- kings$`Start of reign`

endReign <- !is.na(end_of_reign)
startReign <- !is.na(start_of_reign)

# And now I can create a new coloumn called `duration` in the kings dataset:

kings %>%  
  filter(!is.na(endReign)) %>% 
  filter(!is.na(startReign)) %>% 
  mutate(duration = end_of_reign-start_of_reign)

kings_duration <- kings %>%  
  filter(!is.na(endReign)) %>% 
  filter(!is.na(startReign)) %>% 
  mutate(duration = end_of_reign-start_of_reign)


```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE
# The prior task has now enabled me to see/have the data of each monarchs duration of reign. So now I can calculate the average duration of reign for all rulers.
library(tidyverse)

kings_duration %>%  
  filter(!is.na(endReign)) %>% 
  filter(!is.na(startReign)) %>% 
  mutate(duration = end_of_reign-start_of_reign) %>% 
  summarize(mean_duration = mean(duration))

mean(kings_duration$duration, na.rm = TRUE)

```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
# YOUR CODE
# With this code I find every monarch which reign more than the average
kings_duration %>% 
  filter(kings_duration$duration>19.94444)

# Now I have to count all the monarchs which reigned above average
count(kings_duration %>% 
  filter(kings_duration$duration>19.94444))

```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE

# With this code I arrange all monarchs with the longest reign at the top.
kings_duration %>% 
  arrange(desc(kings_duration$duration))

# Now I can find the three monarchs with the longest reign.
kings_duration %>% 
  arrange(desc(kings_duration$duration)) %>% 
  select(Name, duration) %>% 
  slice(1:3)

# Now I will calculate the amount of days they ruled by creating a new coloumn called "days".
kings_duration %>% 
  arrange(desc(kings_duration$duration)) %>% 
  select(Name, duration) %>% 
  slice(1:3) %>% 
  mutate(days = 365*duration)

# And if we want to take the transition year (or leap year) into consideration that happens every fourth year, we can just adjust a little in the number of days for a year.
kings_duration %>% 
  arrange(desc(kings_duration$duration)) %>% 
  select(Name, duration) %>% 
  slice(1:3) %>% 
  mutate(days = 365.25*duration)

```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE

kings_duration %>% 
  mutate(midyear = start_of_reign + (end_of_reign-start_of_reign)/2) %>% 
  ggplot(aes(x = midyear, y = duration)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Duration of Danish Monarchs' Reigns Over Time",
    x = "Midyear of Reign",
    y = "Duration of Reign (Years)"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))

```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


